title: "Module 1 - workshop 3"
author: "Ben Barber"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

#------------------------------------#
# MB5370: Techniques in Marine Science 1
# Programming Fundamentals
 

#------------------------------------#
# Workshop 03. Reproducible research in marine science #

Workshop 3 focused on reproducible research and version control in marine science. During this workshop, I connected RStudio, Git, and GitHub so that my local project could be tracked, backed up, and synced with a remote GitHub repository. I also created an organised project folder structure, set up a Quarto document to work in, and added a README file to explain the purpose and structure of my repository. Through practising the basic Git workflow of saving, staging, committing, pulling, and pushing changes, I developed a clearer understanding of how version control supports transparent, organised, and reproducible scientific coding workflows.

## Running a script in markdown ##
```{r}
source("module_1.R")

getwd()
```

