---
title: "Intro to programming - module 1"
author: "Ben Barber"
date: "2026-05-20"
output: html_document
---
#Introduction
This document contains code and output from module 1 introduction to programming. IT is a working progress and will be upadeted as we go through the workshop

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setting directory #
```{r}
getwd()
```


# Workshop 1 #

```{r Intro to R}
## intro exercises ##

2+1

1:30

6* +2

# Variables and Assignment ####
# This section focuses on understanding how data is stored in R and why that matters

age <- 22
first_name <- 'ben'


age+1

age+ age ##44

#naming object require specifics - can't use numbers, special characters or spaces

##Exercisev <- Use R to calculate the sum of 15, 25.1 and 20.25.

result <- 15 + 25.1 + 20.25 #result<- 60.35

second_result <- 12 + 6 + 84 

result + second_result # = 162.35

##Fucntions in R

years_old <- 57.7

round(years_old) #round up

floor(years_old) #round down

years_old <- 25.765
round (years_olld, 2) # comma after the object to specify argument

##question about functions use args(xxx) and ?round # go to help in console


## Exercsie 2 ##

ben_age <- 22
paste(first_name, 'is', age, "years old")  #Ben is 22 years old


## Debugging ##

# What will this code do?
p <- 2
z <- 5
out <- p * z  # What should the value of out be? <-- will multiple p and z together 
print (out) # What is the value of out? Is it the same as expected?

# exercise #
#my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro"
             "cinco")     # comment: there are is a missing comma in the my_quiz string
#print (my_quis)        # spelling mistake in print function
#str(my_quiz)
#len(my_quiz)       # functoin not real, real functoin = length

#testing

is.character(my_quiz) #true
is.numeric(my_quiz)  #false
#how it shoudl look:

my_quiz <- c("uno",
             "dos",
             "tres",
             "cuatro",
             "cinco")
print(my_quiz )
str(my_quiz)
length(my_quiz)  


## 1.3.11	Data types ##

# character    
# numeric
# integer
# logical
# complex
# raw


#class() - what kind of object is it (high-level)?
#  typeof() - what is the object's data type?
#length() - how long is it?
#attributes() - does it have any metadata? 


## Exercise ##
my_numeric_variable  <-  6.3
my_character_varaible <- "ben"
my_integer_varaible <- 8
my_logical_varaible <- 9 < 12

typeof(my_numeric_variable)
typeof(my_character_varaible)
typeof(my_integer_varaible)
typeof(my_logical_varaible)

class(my_numeric_variable) #numeric
class(my_character_varaible) #character
class(my_integer_varaible) #numeric
class(my_logical_varaible) #logical



## 1.3.12 Data structures ##
# R has many data structures. These include:
#  atomic vector                             vector is most common
#  list
#  matrix
#  data frame
#  factors

y <- c(1, 2, 3) #creating a numeric vector 

z <- c("Sarah", "Tracy", "Jon")   #crating character vector

class(y) #numeric
class(z) #character        type within a vector must be the same 

# Lists : list can be of several different types.
x <- list(1, "a", TRUE)
x

x[[2]]   # retrieve individual elements using double square brackets to reference their index.


## 1.3.13	Data frames and tibbles ##

#dataframe is the R version of a spreadsheet

my_dataframe <- data.frame (no = c(1,2,3), c("Tracey", "John", "Pete"), c(TRUE, FALSE, TRUE))
                           
my_dataframe

str(my_dataframe)

# checking the dataframe
my_dataframe$no = as.factor(my_dataframe$no)   #A factor is a categorical type, so here we are telling R 
str (my_dataframe)                            #that column 1 is actually storing a category (a group membership, 
                                               #say a numeric code for which team each of these people are in) 
                                              #rather than a real continuous number. 

#can use NA as a cell holder when we don't have the data <-  can use na.omit function 


## 1.3.14	Packages and libraries ##

#packages are a collection of functions

#install.packages("tidyverse") # download and install
library("tidyverse") # load into current workspace

#install.packages ("ggplot2") # comment out so it doesn’t run again, once its installed on computer it doesn't need to be reinstalled


#install.packages("earthtones")
#install.packages ("redlistr")
#install.packages("terra")

library (ggplot2)


## 1.3.15	Coding best practices ##

# title
# Read data .csv
# load packages
```


# Workshop 2 #

```{r visulisation - plotting}
# load package

library(tidyverse)

#explore ggplot dataframe
mpg
?mpg
```

##first plot - ggplot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```
geom_point() adds a layer of points to your plot,

ggplot plot template:
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

## empty plot
```{r}
ggplot() 
```
##plot with data argument
```{r}
ggplot(data = mpg)
```
plots require 2 attributes: Geom and aesthetics

##plot with colour arguments - class to colour
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```

##change class to size
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
#> Warning: Using size for a discrete variable is not advised. 
```

##Change point transparency by class (alpha)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

## Change point shape by class.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

## change to specifc colour - blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") 
#create all dot points as blue
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 6))
```

##example error:
```{r}
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy)) 
# the + should be on top line
```

## Facet and panel plots
To facet your plot using a single variable, use facet_wrap(). Facet wrap syntax is in the function of a formula (kind of like a linear model formula), where the ~ dictates which variable you want to subset your data with. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

##include more than 1 variable 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

## Period makes varaible not be incldued twice 
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

nrow and ncol refer to the number of rows or columns, eg.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2, ncol = 8)
```

## Fitting simple lines
To display data as points, use geom_point().
= ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
  
To display the same data as a smooth line fit through the points use geom_smooth().
= ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
  
##combining both togehter:
```{r}
ggplot(data = mpg) + 
 geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

## Incorporating colour and points 
```{r}
ggplot(data = mpg) + 
   geom_point(mapping = aes(x = displ, y = hwy, colour = drv, shape = class))+
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, colour = drv))
```

##grouping arguements
can use group argument to show that our is grouped, wont add legend but can easlity be used to cary plots by the discrete vairables 

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

changing the colour and incorporating both point and smooth:
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,
  ) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = drv, shape = class))
```

do it this way duplicate the x and y arguements but ggplot allows for the overarching mappings:

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```

##combining together
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

can include subsets under grom_smooth:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

##Exercise:
What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_line(mapping = aes(color = class))
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_boxplot(mapping = aes(color = class))
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ)) + 
  geom_histogram(mapping = aes(color = class)) #histogram only has one varaible
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_area(mapping = aes(color = class))
```

##excerise
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

these produce the same graphs as the agrument have been specifcied earlier in the code

##Transformations and Stats - using diamonds dataset
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```
bar charts, histograms, and frequency polygons “bin” your data and then plot bin counts (the number of points that fall in each bin).

smoothers fit a model to your data and then plot predictions from the model.

boxplots compute a robust summary of the distribution and then display a specially formatted box.

You can generally use geoms and stats interchangeably. For example, you can recreate the previous plot using stat_count() instead of geom_bar().
```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

##Overriding defaults 
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

overriding - showing proportion rather than counts 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
#> Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
```

##Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## Aesthetic adjustments
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) # this colours the outline of the bar
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) #this fills the bar with colour
```
coloured by clarity 
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
 position adjustments is vital, it allows you to customise your plots in three ways, identity (raw data), fill (changes heights) and dodge (which forces ggplot2 to not put things on top of each other)
 
```{r}
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```
 
position = "fill" works like stacking, but makes each set of stacked bars the same height.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

position = "dodge" places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

##The layered grammar of graphics

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <FACET_FUNCTION>






#-----Workshop 3------#


Running a script in markdown:
```{r}
source("module_1.R")

getwd()
```
The workshop was spent linking our R studio, Git and Github. I have now linked my local work to my repo on github.



#-----Workshop 4------#

Exercise: AI-Driven Data Visualisation Challenge

```{r}
# Load tidyverse and palmerpenguins libraries

library(tidyverse)
library(palmerpenguins)
#install.packages("palmerpenguins")

```
#Instructing copliot

```{r}
# Create a ggplot scatter plot using the penguins dataset. 
# Plot bill_length_mm on the x-axis and bill_depth_mm on the y-axis.
# Color points by species and use geom_point with size 3.
# Add a minimal theme and clean labels.

ggplot(data = penguins) + 
  geom_point(mapping = aes(x = bill_length_mm, y = bill_depth_mm, color = species), size = 3) +
  theme_minimal() +
  labs(title = "Penguin Bill Dimensions",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species")
```

#Open-Ended Comparative Plotting

Option A (Distribution Check):

```{r}
#create a ggplot boxplot using the penguins dataset.
#plot flipper_length_mm on the y-axis and species on the x-axis.
#facet the plot by island and use a custom color palette for the species.
#add a title and clean labels.
ggplot(data = penguins) + 
  geom_boxplot(mapping = aes(x = species, y = flipper_length_mm, fill = species)) +
  facet_wrap(~ island) +
  scale_fill_manual(values = c("Adelie" = "blue", "Chinstrap" = "green", "Gentoo" = "orange")) +
  labs(title = "Flipper Length by Species and Island",
       x = "Species",
       y = "Flipper Length (mm)",
       fill = "Species") +
  theme_minimal()
```


Option B (Biomass Correlation):
```{r}
#Instruct Copilot to build a scatter plot mapping body_mass_g against flipper_length_mm, adding a linear regression trend line (geom_smooth) for each species.

#create a ggplot scatter plot using the penguins dataset.
#plot body_mass_g on the y-axis and flipper_length_mm on the x-axis.
#add a linear regression trend line (geom_smooth) for each species.
ggplot(data = penguins) + 
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_smooth(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species), method = "lm") +
  labs(title = "Body Mass vs Flipper Length by Species",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Species") +
  theme_minimal()



```


##^^^^ finish


#chattr (Conversational Workspace Engine)

```{r}
#install.packages("ellmer")
#install.packages("chattr")

```

Configuring chattr to model 

```{r}
#usethis::edit_r_profile()

# Option A: For Google Gemini
#options(.chattr_chat = ellmer::chat_google_gemini())

```
using the ai chat box, Exercise: Prompting for Context:
```{R}
#library(ggplot2)

data(iris)

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()

```
#Conditional Logic - filtering your data
example:
```{r}
# Simple Example: Checking Sea Surface Temperature (SST)
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}

```

Base R: if and else
The classic base R conditional structure screens a single value at a time. Its basic anatomy uses round brackets () for the logical condition and curly braces {} for the action if that condition is true.


Tidyverse Alternative: Vectorised Conditionals (dplyr):
if_else()
The if_else(condition, true_output, false_output) function evaluates an entire column item-by-item:

```{r}
# Load tidyverse (if not already loaded)
library(tidyverse)

# Sample coral data
coral_monitoring <- tibble(
  site = c("Site_A", "Site_B", "Site_C"),
  depth_m = c(12, 35, 8)
)

# Classify depth using if_else
coral_monitoring <- coral_monitoring %>% 
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef"))

```
what do you think that function mutate()  does? 
>The mutate() function from the dplyr package (part of the tidyverse) adds new columns to a data frame, or modifies existing ones, based on functions of existing columns. It's a powerful tool for creating new variables or transforming existing ones.


What about the %>% symbol? Use your developing programmer skills to find out more information about these two functions and record your findings now in your qmd.

>The %>% symbol, known as the pipe operator (or forward pipe operator), is a key feature from the magrittr package, which is automatically loaded when you load dplyr or the tidyverse.
Its primary purpose is to make code more readable and easier to write by chaining operations together. Instead of nesting function calls from the inside out, the pipe allows you to read code from left to right, passing the result of one function as the first argument to the next function.
Consider the following:

function_c(function_b(function_a(data, arg1), arg2), arg3)

With the pipe operator, this becomes:
  
data %>%
  function_a(arg1) %>%
  function_b(arg2) %>%
  function_c(arg3)

This structure often makes the sequence of operations much clearer to follow.



case_when()
When you have more than two possibilities, stacking multiple if-else blocks becomes unreadable. case_when() provides a clean, sequential evaluation layout using formula tildes (~):

```{r}
coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))

```


#Task:  Environmental Stress Classification

```{r}
library(dplyr)
library(tibble)

marine_stations <- tibble(
  salinity = c(35, 28, 32, 12)
) %>%
  mutate(
    environment_type = case_when(
      salinity < 15 ~ "Estuarine",
      salinity >= 15 & salinity <= 30 ~ "Brackish",
      salinity > 30 ~ "Marine"
    )
  )

marine_stations
```

#Automation — Iterating Efficiently

Base R: for-loops

```{r}
#A for-loop repeats a code chunk for each element in a designated sequence.
# Classic loop anatomy
for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}


```
If we want to track real-world data, loops can iterate over file directories (e.g., reading 12 monthly turtle tracking CSVs sequentially):

```{r}
# Loop across index sequences
transect_lengths <- c(50, 100, 25, 75)

for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}
```
#Tidyverse Alternative: Functional Programming (purrr)
Function     Expected Output Data Type
map()        Returns a flexible List structure
map_dbl()    Returns a vector of Decimals / Numeric values
map_chr()    Returns a vector of Text strings / Characters
map_df()     Returns a combined Data Frame / Tibble

Comparison Example: Calculating Square Roots of Site Vectors

Using a standard for-loop:
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])

```{r}
#Using a standard for-loop:
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])}
```

```{r}
#Using the purrr equivalent:
library(purrr)
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)
```


You can combine purrr pipelines seamlessly inside data frames to split analysis across separate groups:

```{r}
# Iterate a summary function over split lists of data
iris %>% 
  split(.$Species) %>% 
  map(~summary(.x))
```
#Task using chattr: 
Use chattr to convert a complex operation. Prompt the engine: "I have a list of three vectors containing fish count data. Write a for-loop to calculate the mean of each, and then show me the exact parallel way to do it using purrr's map_dbl function." Compare the visual simplicity.
```{r}
library(purrr)
library(dplyr)

# Create the list of vectors
fish_data <- list(
  vector1 = c(10, 12, 15, 11, 13),
  vector2 = c(5, 7, 6, 8, 9),
  vector3 = c(20, 22, 19, 21, 23)
)

# Using a for-loop
means_for_loop <- numeric(length(fish_data))
for (i in seq_along(fish_data)) {
  means_for_loop[i] <- mean(fish_data[[i]])
}

# Using purrr::map_dbl
means_map_dbl <- map_dbl(fish_data, mean)
```

```{r}
print(means_for_loop)
print(means_map_dbl)
```

#Writing custom functions

Function Construction Blueprint

```{r}
# Function Definition
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
}

# Utilizing your custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84)

```


# Student Exercise 4: Custom Function Construction Task

Build a custom function named convert_temp_c_to_f that accepts a single argument representing temperature in Celsius. The function should perform the math conversion: F = (C * 9/5) + 32. Include a conditional safety check using an if statement that halts execution via stop() if the entered value is below absolute zero (-273.15 degrees Celsius). Test your function with both a valid temperature and an invalid value to ensure the error handling behaves as expected.


```{r}
# Define the custom function
convert_temp_c_to_f <- function(celsius_temp) {
  # Conditional safety check for absolute zero
  if (celsius_temp < -273.15) {
    stop("Temperature cannot be below absolute zero (-273.15 degrees Celsius).")
  }

  # Perform the conversion
  fahrenheit_temp <- (celsius_temp * 9 / 5) + 32
  return(fahrenheit_temp)
}

# Test with a valid temperature
valid_celsius <- 25
valid_fahrenheit <- convert_temp_c_to_f(valid_celsius)

# Test with an invalid temperature to ensure error handling
invalid_celsius <- -300

```

```{r}
convert_temp_c_to_f(120) # = 248

convert_temp_c_to_f(-300) #Error in convert_temp_c_to_f(-300) : Temperature cannot be below absolute zero (-273.15 degrees Celsius).
```


challenage 1

In real-world field surveys, calculations often depend on matching values across multiple lists or matrices. Read the documentation for purrr::map2_dbl(). Given a vector of raw survey counts counts <- c(15, 24, 8, 42) and a vector of species-specific scaling factors factors <- c(1.2, 0.8, 2.5, 1.1), write a mapping statement that iterates across both vectors simultaneously to return a single vector of scaled abundance values. Do not use a manual index loop.

```{r}
library(purrr)

# Define the vectors
counts <- c(15, 24, 8, 42)
factors <- c(1.2, 0.8, 2.5, 1.1)

# Use map2_dbl to iterate across both vectors simultaneously
scaled_abundance <- map2_dbl(counts, factors, ~ .x * .y)

```


Challenge 2: The Robust Data Wrangling Pipeline Function

```{r}
library(dplyr)
library(rlang) # Needed for !!sym() to handle string column names elegantly

# Define the custom function
clean_and_classify_survey <- function(df, col_to_check, threshold_value) {
  # 1. Safely check if the specified column name exists
  if (!col_to_check %in% names(df)) {
    stop(paste0("Error: Column '", col_to_check, "' not found in the data frame."))
  }

  # 2. Filter out rows with missing values (NA) in the specified column
  # 3. Use pipe operator and mutate() along with a vectorized conditional
  cleaned_df <- df %>%
    filter(!is.na(!!sym(col_to_check))) %>% # !!sym() allows using a string as a column name
    mutate(
      classification = if_else(
        !!sym(col_to_check) > threshold_value,
        "Above Threshold",
        "Below Threshold"
      )
    )

  # 4. Return the cleaned, altered data frame
  return(cleaned_df)
}

# --- Test your custom pipeline ---

# Create a mock tibble with at least one missing value
mock_survey_data <- tibble(
  site = c("A", "B", "C", "D", "E"),
  species = c("Fish1", "Fish2", "Fish1", "Fish3", "Fish2"),
  count = c(10, NA, 25, 7, 18),
  biomass = c(100, 120, 150, NA, 110)
)

# Define a threshold value
my_threshold <- 15

# Test with valid inputs
cat("--- Testing with valid inputs ---\n")
cleaned_classified_data <- tryCatch(
  {
    clean_and_classify_survey(mock_survey_data, "count", my_threshold)
  },
  error = function(e) {
    message("An error occurred: ", e$message)
    NULL
  }
)

# Show the result if successful
if (!is.null(cleaned_classified_data)) {
  print(cleaned_classified_data)
}

# Test with an invalid column name to confirm error handling
cat("\n--- Testing with an invalid column name ---\n")
tryCatch(
  {
    clean_and_classify_survey(mock_survey_data, "non_existent_col", my_threshold)
  },
  error = function(e) {
    message("Successfully caught error: ", e$message)
  }
)

```